---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

To set up necessary packages and libraries:
```{r echo=TRUE,message=FALSE}
  #to install packages if it is required
  if (!require("dplyr")){
    install.packages("dplyr")
  }
  if (!require("ggplot2")){
    install.packages("ggplot2")
  }

  if (!require("lubridate")){
    install.packages("lubridate")
  }
  if (!require("lattice")){
    install.packages("lattice")
  }

  #to install libraries if it is required
  if (!require(dplyr)){
    library(dplyr)  
  }

  if (!require(ggplot2)){
    library(ggplot2)  
  }  
  if (!require(lubridate)){
    library(lubridate)
  }
  if (!require(lattice)){
    library(lattice)
  }


```

###**Copy data source file into work directory!!!**#####

## Loading and preprocessing the data

The data loading is with: read.csv() function.
To change data into tbl format to analyze further:

```{r}
setwd("D:/Dimon/Coursera/Reproducible_Research/Project1/to_Git")
data<-read.csv("activity.csv")
tbl_data<-tbl_df(data)
rm(data)
head(tbl_data)
```


## What is mean total number of steps taken per day?

- to remove NA from the subset
```{r}
rm_na<-filter(tbl_data,!is.na(steps))
```
- group data, summarize and make the histogram:
```{r}
by_date<-group_by(rm_na,date) 
total<-summarize(by_date,total=sum(steps))
with(total,hist(total,10,col=date
                ,main="Total step frequancy per one day"
                ,xlab="The total steps per one day"))

```

- calculate mean and median:

```{r}
mean<-mean(total$total)
median<-median(total$total)
```

Mean is `r mean`.    
Median is `r median`.

## What is the average daily activity pattern?
* group data by interval
* calculate mean by interval
* draw a plot
```{r}
intervals<-group_by(rm_na,interval)
mn_intervals<-summarize(intervals,steps=mean(steps))


g<-ggplot(data=mn_intervals,aes(x=interval,y=steps))+geom_line()+ xlab("Interval")
g<-g+ylab("Step mean q-ty ")
g
```

calculate the maximum number of steps:
```{r}
max_interval<-mn_intervals[which(mn_intervals$steps==max(mn_intervals$steps)),]$interval
```

the interval with the maximum number of steps is from `r max_interval` and `r max_interval+5`.  
the max step q-ty is `r max(mn_intervals$steps)`

## Imputing missing values

calculate missing values:
```{r}
na_steps<-is.na(tbl_data$steps)
na_qty<-sum(na_steps)
```
Total numbe of missing values is `r na_qty`

* Join mean interval steps across all days to the original dat set tbl_data:
```{r}
join_tbl<-left_join(tbl_data,mn_intervals,by="interval")
```

* Assigning the missing stpes field with the mean steps of that interval
```{r}
join_tbl[na_steps,]$steps.x<-join_tbl[na_steps,]$steps.y
```

* Selecting the needed data set from the joined data set
```{r}
filled_df<-select(join_tbl,steps=steps.x,date,interval)
```

* grouping the selected data set by date
```{r}
filled_by_date<-group_by(filled_df,date)
```
* New total stpes taken per day
```{r}
filled_total<-summarize(filled_by_date,total=sum(steps))
with(filled_total,hist(total,10,col=date,
                main="Frequency of the total steps in one day",
                xlab="The total steps in one day"))
```

* New average and median total steps taken per day
```{r}
filled_mean<-summarize(filled_total,mean(total))
filled_median<-summarize(filled_total,quantile(total,probs=0.5))
```
After filling the NA records, the new mean total steps taken per day is `r filled_mean`.  
The new median is `r filled_median`.  

```{r}

par(mfrow = c(1, 2))
hist(total$total, breaks = 10, xlab = "Steps", main = "Initial Histogram", ylim = c(0, 20))
hist(filled_total$total, breaks = 10, xlab = "Steps", main = "NA removed Histogram", ylim = c(0, 20))

```


The new data set differes from the original one (with missing values).Application of avg numbers leads to a higher distribution (look at the 6th interval).  
Besides averages are the same but medians differ. On the whole, median and avg in both data sets are almost the same.


## Are there differences in activity patterns between weekdays and weekends?

* add new factor coloumn (weekends, weekdays)
```{r}
filled_df<-mutate(filled_df, weekend=wday(as.Date(filled_df$date), label=TRUE) %in% c('Sat','Sun'))
```

* group the new filled data frame by weekend and interval
```{r}
filled_by_int_wk<-group_by(filled_df,weekend,interval)
```
* Calcating the average interval mean based on groups
```{r}
filled_mn_int<-summarize(filled_by_int_wk,steps=mean(steps))

m<-filled_mn_int$weekend==TRUE
index<-which(m %in% c(TRUE))
filled_mn_int$weekend[index]<-"Weekends"

m<-filled_mn_int$weekend==FALSE
index<-which(m %in% c(TRUE))
filled_mn_int$weekend[index]<-"Weekdays"

```
* Plotting the averge interval steps taken per day for the weekday and weekend individually
```{r}

xyplot(steps ~ interval | weekend, filled_mn_int, type = "l", layout = c(1, 2), ylab = "Number of Steps", xlab = "Interval", main = "Time Series for Weekend and Weekday Activity Patterns")
```
